Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-05-17 paper/main.tex
atct

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}

@@

\section{Data And Visualizations}
\label{sec:atct}
\note{why why why?}
We use concepts from algberaic topolology and category theory to formally express the structure visualization components are expected to preserve.  Fiber bundles and sheaves, which come from algebraic topolology, provide a uniform generalizable way to describe data and visualizations. Category theory is a method to describe how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose categorical constructions of visualization components as specifications for implementable components.

@@

\begin{figure}[h!]
  \label{fig:artist:equivariance}
  \includegraphics[width=1\columnwidth]{equivariance.png}
  \caption{}
\end{figure}
For example, in \autoref{fig:artist:equivarance}, the measurable variable is color. This is a visual representation of the data shown in \autoref{fig:atct:phi}, and as such the equivariant transformations are an equivalent transposition and scaling of the colors. This visualization is equivariant with respect to base space transformations, as defined in \autoref{eq:atct:equivarance:verify:base}, because the color values at the new position at the old position $measure_\dbasepoint^{\prime} = \measure_{\dbasepoint}$. This visualization is also equivariant with respect to fiber wise transformations, as defined in \autoref{eq:atct:equivarance:verify:fiber}, because the colors are consistently scaled in the same was the data. For example, the values that have become 2 and 4 in the $\dfunctc$ and $\dfunc$ panels are colored the same as the original 2 and 4 values in the first panel. The equivariance in this visualization is composable, as shown in the colors being both transposed and scaled correctly in the $\dfunc$ panel.

@@

The circle-line glyph in \autoref{fig:artist:operator} can be generated via a line plot and highlighted points in that plot, which is the addition case of two datasets with the same fields and different continuities. The same circle-line glyph can also be generated via the multiplication case as different fields of the same bundle. As mentioned in \autoref{sec:artist:operator:multiplication}, multiplication is implemented as addition of two copies of a base space. Although we have been discussing the trivial cases of adding observations or adding fields, this merging of artists in datasets can be generalized:
\begin{equation}
  \vartistc(\cgamma{\bigsqcup_{i} \dbasec^{i}}{\bigoplus_{i}\dtotalc^{i}}) \coloneqq \sum_{i}
  \vartisct_{i}(\cgamma{\dbasec^{i}}{\dtotalc^{i}}) 
\end{equation}

@@

An encoding function $\vchannelc$ is equivariant if the change in data, as defined in \autoref{sec:atct:sheaves:measurments}, and change in visual components are equivariant. Since $\dtotal$ and $\vtotal$ are over the same base space and are pointwise, the base space change $\dfunch_{\dtotal}$ applies to both sides of the equation 
\begin{equation}
  \vchannelc(\dsectionc_{\dtotalc}(\dfunchc_{\dbasec}(\dbasepointc^{\prime}))) = \vsectionc(\dfunchc_{\dbasec}(\dbasepointc^{\prime}))
\end{equation}
and therefore there should not be a change in encoding. On the other hand, a change in the data values $\dfunct_{\dtotal}$ must have an equivalent change in visual components
\begin{equation}
  \dfunctc_{\vtotal} \vchannelc(\dsectionc(\dbasepoint)) = \vchannelc(\dfunctc_{\dtotalc}(\dsectionc(\dbasepointc)))
\end{equation}
The change in visual components $\dfunct_{\vtotal}$ is dependent both on $\dfunct_{\dtotal}$ and the choice of visual encoding. As mentioned in \autoref{sec:related-work:equivariance}, this is why Bertin and many others since have advocated choosing an encoding that has a structure that matches the data structure\cite{bertinSemiologyGraphicsDiagrams2011a}. For example choosing a quantative colormap to encode quantative data if the $\dfunctc$ operation is scaling, as in \autoref{fig:artist:equivariance}.

\subsection{Composite Measurable Components Into Visual Elements}
As shown in \autoref{fig:construction:flow-chart}, the compositor function $\vmarkc$ transforms the measurable components into properties of a visual element. The compositing function $\vmarkc$ transforms the sections of visual elements $\vsectionc$ into sections of graphics $\gsectionc$.
\begin{equation}
  \vmarkc: \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}} \rightarrow \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}}
\end{equation}

\begin{equation}
  \label{eq:construction:q:fabrication}
  \begin{tikzcd}
      \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}} 
      \arrow[rr, "\vchannel", color=artist] 
      \arrow[rrrr, "\vmarkc^{\prime}"', bend right, color=artist, dashed] &  & \cgamma{\opensetc}{\vtotalc^{\prime}\restriction_{\opensetc}} 
      \arrow[rr, "\vmarkc", color=artist] &  & \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}}
      \end{tikzcd} 
\end{equation}

>>>>

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}
\note{please prioritize flagging content that needs major revisions: (1) Wrong (2) Missing (3) Horrific  (4) Incoherant}

@@

\section{Data And Visualizations}
\label{sec:atct}
A goal of this work is to develop guidelines for building composable structure preserving functions for translating data into visual representations. Concepts in algebraic topology and category theory provide a language for formally expressing the structure that these visualizations are expected to preserve. Fiber bundles and sheaves, which come from algebraic topolology, provide a uniform generalizable way to describe data and visualization. Category theory is a method to describe how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose categorical constructions of visualization components as specifications for implementable components.

@@

\begin{figure}[h!]
  \label{fig:artist:equivariance}
  \includegraphics[width=1\columnwidth]{equivariance.png}
  \caption{This artist is equivariant because the input data and output graphic representation are transposed $\dfunch$, scaled $\dfunct$ and both transposed and scaled $\dfunc$ in the same manner.}
\end{figure}
For example, in \autoref{fig:artist:equivarance}, the measurable variable is color. This is a visual representation of the data shown in \autoref{fig:atct:phi}, and as such the equivariant transformations are an equivalent transposition and scaling of the colors. This visualization is equivariant with respect to base space transformations, as defined in \autoref{eq:atct:equivarance:verify:base}, because the color values at the new position at the old position $measure_\dbasepoint^{\prime} = \measure_{\dbasepoint}$. This visualization is also equivariant with respect to fiber wise transformations, as defined in \autoref{eq:atct:equivarance:verify:fiber}, because the colors are consistently scaled in the same was the data. For example, the values that have become 2 and 4 in the $\dfunctc$ and $\dfunc$ panels are colored the same as the original 2 and 4 values in the first panel. The equivariance in this visualization is composable, as shown in the colors being both transposed and scaled correctly in the $\dfunc$ panel.

@@

The circle-line glyph in \autoref{fig:artist:operator} can be generated via a line plot and highlighted points in that plot, which is the addition case of two datasets with the same fields and different continuities. The same circle-line glyph can also be generated via the multiplication case as different fields of the same bundle. As mentioned in \autoref{sec:artist:operator:multiplication}, multiplication is implemented as addition of two copies of a base space. Although we have been discussing the trivial cases of adding observations or adding fields, this merging of artists in datasets can be generalized:
\begin{equation}
  \vartistc(\cgamma{\bigsqcup_{i} \dbasec^{i}}{\bigoplus_{i}\dtotalc^{i}}) \coloneqq \sum_{i}
  \vartistc_{i}(\cgamma{\dbasec^{i}}{\dtotalc^{i}}) 
\end{equation}

@@

An encoding function $\vchannelc$ is equivariant if the change in data, as defined in \autoref{sec:atct:sheaves:measurments}, and change in visual components are equivariant. Since $\dtotal$ and $\vtotal$ are over the same base space and are pointwise, the base space change $\dfunch_{\dtotal}$ applies to both sides of the equation 
\begin{equation}
  \vchannelc(\dsectionc_{\dtotalc}(\dfunchc_{\dbasec}(\dbasepointc^{\prime}))) = \vsectionc(\dfunchc_{\dbasec}(\dbasepointc^{\prime}))
\end{equation}
and therefore there should not be a change in encoding. On the other hand, a change in the data values $\dfunct_{\dtotal}$ must have an equivalent change in visual components
\begin{equation}
  \dfunctc_{\vtotalc} \vchannelc(\dsectionc(\dbasepoint)) = \vchannelc(\dfunctc_{\dtotalc}(\dsectionc(\dbasepointc)))
\end{equation}
The change in visual components $\dfunct_{\vtotal}$ is dependent both on $\dfunct_{\dtotal}$ and the choice of visual encoding. As mentioned in \autoref{sec:related-work:equivariance}, this is why Bertin and many others since have advocated choosing an encoding that has a structure that matches the data structure\cite{bertinSemiologyGraphicsDiagrams2011a}. For example choosing a quantative colormap to encode quantative data if the $\dfunctc$ operation is scaling, as in \autoref{fig:artist:equivariance}.

\subsection{Composite Measurable Components Into Visual Elements}
As shown in \autoref{fig:construction:flow-chart}, the compositor function $\vmarkc$ transforms the measurable components into properties of a visual element. The compositing function $\vmarkc$ transforms the sections of visual elements $\vsectionc$ into sections of graphics $\gsectionc$.
\begin{equation}
  \vmarkc: \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}} \rightarrow \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}}
\end{equation}
The compositing function is map from sheaves over $\dbase$ to sheaves over $\gbase$. This is because, as described in \autoref{fig:construction:xi}, the graphic section must be evaluated on all points in the graphic space to generate the visual element corresponding to a data record at a single point $\vartist(\dsection(\dbasepoint)) = \gsection(\vindexpre(\dbasepoint))$.

Since encoder functions are infinitely composable, as described in \autoref{eq:construction:nu:fabrication}, a new compositor function $\vmarkc$ can be constructed by precomposing $\vchannelc$ functions with the existing $\vmarkc$.

\begin{equation}
  \label{eq:construction:q:fabrication}
  \begin{tikzcd}
      \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}} 
      \arrow[rr, "\vchannel", color=artist] 
      \arrow[rrrr, "\vmarkc^{\prime}"', bend right, color=artist, dashed] &  & \cgamma{\opensetc}{\vtotalc^{\prime}\restriction_{\opensetc}} 
      \arrow[rr, "\vmarkc", color=artist] &  & \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}}
      \end{tikzcd} 
\end{equation}
The composition in \autoref{eq:construction:q:fabrication} means that different measurable components can yield the same visual elements.

\begin{figure}[h!]
  \label{fig:construction:q}
  \includegraphics[width=1\columnwidth]{q.png}
  \caption{This simple $\vmarkc$ assembles a circular visual element that is the color specified in $\vsection(\dbasepoint)$ and is at the intersection specified in $\vsection(\dbasepoint)$}
\end{figure}
As shown in \autoref{fig:construction:q}, a set of  $\vchannel$ functions individually convert the values in the data record to visual components. Then the $\vmarkc$ function combines these visual encodings to produce a graphic section \gsection. When this section is evaluated on the graphic space associated with the data $\gsection(\vindexpre(\dbasepoint))$, it produces a blue circular marker at the intersection of the x and y positions listed in \vsection. The composition rule in \autoref{eq:construction:q:fabrication} means that developers can implement $\vmarkc$ as drawing circles or can implement a $\vmarkc$ that draws arbitrary shapes, and then provide different $\vchannelc$ adapters, such as one that specifies that the shape is a circle.
